---
title: "Exploratory Data Analysis"
author: "Mingye Xie, Yujing Fu"
output: 
  html_document:
    code_folding: hide
always_allow_html: true
---

```{r setup, message = FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(rvest)
library(janitor)
library(geosphere)
library(sf)
if (!requireNamespace("leaflet", quietly = TRUE)) install.packages("leaflet")
library(leaflet)

knitr::opts_chunk$set(
  warning = FALSE,   # Suppresses warnings
  message = FALSE    # Suppresses messages
)

```

```{r, include=FALSE}
# data import 
rat_df = read.csv("data/rat_2020_2024.csv")
food_scrap_df = 
  read.csv("data/food_scrap_drop_off.csv")
zhvi_df = read.csv("data/ny_zhvi_df_2020_2024.csv")
zori_df = read.csv("data/ny_zori_df_2020_2024.csv")
nyc_boroughs = st_read("Borough Boundaries.geojson")
```

# Temporal patterns of rodent activity
## Is rodent activity differs by Month? 
```{r}
# Filter for rows where the result indicates rodent activity
data_rodent <- rat_df %>% filter(grepl("Rat Activity", result, ignore.case = TRUE))

# Group by neighborhood (NTA), year, and month to count occurrences of rodent activity
rodent_activity_summary <- data_rodent %>% 
  group_by(nta, year, month) %>% 
  summarise(count = n(), .groups = 'drop') %>% 
  arrange(desc(count))

# Print the top neighborhoods with the highest rodent activity
print(head(rodent_activity_summary))
```

```{r}
# Filter data for the year 2020
rodent_activity_2020 <- rodent_activity_summary %>% filter(year == 2020)

# Create a bar plot for rodent activity in 2020
ggplot(data = rodent_activity_2020, aes(x = factor(month, levels = 1:12), y = count)) +
  geom_bar(stat = "identity", position = "dodge", fill = "steelblue") +
  labs(title = "Rodent Activity in 2020",
       x = "month",
       y = "Rodent Activity Count") 

```

This bar plot shows how rodent activity varies across months in 2020. This helps us to identify seasonal trends, such as higher rodent activity in summer months. We found that there is a clear peak in rodent activity during the summer months (July and August). A significant drop in activity is observed in the spring (April). Understanding monthly trends in rodent activity can inform targeted pest control strategies during peak seasons.

## How does rodent activity fluctuate monthly across the years?
```{r}
# Aggregate data by year and month
yearly_monthly_summary <- data_rodent %>%
  group_by(year, month) %>%
  summarise(total_activity = n(), .groups = 'drop') %>%
  arrange(year, month)

# Create a line plot showing rodent activity over time
ggplot(yearly_monthly_summary, aes(x = as.numeric(month), y = total_activity, color = as.factor(year), group = year)) +
  geom_line(size = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Monthly Rodent Activity by Year (2020-2024)",
       x = "Month",
       y = "Total Rodent Activity Reports",
       color = "Year") +
  theme_minimal() +
  theme(legend.position = "right")
```

We further examined patterns in rodent activity across months and years and observed that activity generally peaks during the summer months (June to September), regardless of the year. However, some years show anomalies, such as a noticeable dip in October 2024.<br>
<br>
Interestingly, 2020 had the lowest rodent activity over the five-year period, likely due to the COVID-19 lockdown. In contrast, 2022 recorded the highest levels of activity.<br>

# Geographical patterns of rodent activity
## What are the distribution patterns of food scrap drop-off sites?
### Distribution of food scrap drop-off sites by borough

```{r}
# Analyze the distribution of food scrap drop-off sites by borough
borough_distribution <- food_scrap_df %>% 
  group_by(borough) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))

# Display the borough distribution
print("Borough Distribution:")
print(borough_distribution)
```

```{r}
# Borough distribution plot
ggplot(borough_distribution, aes(x = reorder(borough, -count), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Distribution of Food Scrap Drop-off Sites by Borough",
       x = "Borough",
       y = "Number of Sites") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

After observing the temporal patterns of rodent activity, we aim to explore its geographic distribution. We are interested in determining if there is a correlation between the presence of food scrap drop-off sites across New York City's boroughs and the levels of rodent activity. To begin this inquiry, we first sought to identify any disparities in access to food scrap sites between the boroughs.<br>
<br>
Our findings indicate that Manhattan has the highest number of drop-off sites, with Brooklyn ranking second. In contrast, the Bronx, Queens, and Staten Island have a significantly lower number of sites. This disparity may indicate potential inequities in access to food scrap recycling programs, with Staten Island residents being particularly affected.

### Distribution of food scrap drop-off sites on map
```{r}
# Analyze the temporal patterns
# First, let's create a simplified season categorization
food_scrap_df <- food_scrap_df %>%
  mutate(
    operation_type = case_when(
      grepl("Year Round", open_month, ignore.case = TRUE) ~ "Year Round",
      grepl("Seasonal|Spring|Summer|Fall|Winter", open_month, ignore.case = TRUE) ~ "Seasonal",
      TRUE ~ "Specific Months"
    )
  )

# Define a color palette for boroughs
pal <- colorFactor(
  palette = c("red", "green", "blue", "orange", "purple"),
  domain = food_scrap_df$borough
)

leaflet() %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  addPolygons(
    data = nyc_boroughs,
    fillColor = "lightgray",
    color = NA,
    weight = 1,
    opacity = 0.8,
    label = ~boro_name
  ) %>%
  addCircleMarkers(
    data = food_scrap_df,
    lng = ~longitude, lat = ~latitude,
    color = ~pal(borough),  # Use the color palette for boroughs
    popup = ~paste("Borough:", borough),
    radius = 4,
    fillOpacity = 0.7
  ) %>%
  addLegend(
    "bottomright",
    colors = c("red", "green", "blue", "orange", "purple"),
    labels = c("Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"),
    title = "Borough"
  )
```

Next, we visualized the clustering and disparities in the availability of drop-off sites. We found that drop-off sites are densely clustered in Manhattan and Brooklyn. The sparse distribution in Staten Island reflects a lower number of food scrap programs in this borough. Queens and the Bronx show a moderate distribution but lack the density observed in Manhattan.

## Geographical distribution of rodent activity and its interaction with food scrap drop-off sites
```{r}
plot_county <- ggplot(zhvi_df, aes(x = CountyName, fill = CountyName)) +
  geom_bar() +
  labs(
    title = "Demographic Differences by County",
    x = "County",
    y = "Count",
    fill = "County"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),  # Rotate and resize x-axis text
    plot.title = element_text(size = 16, face = "bold"),  # Enlarge title
    axis.title.x = element_text(size = 14),  # Enlarge x-axis label
    axis.title.y = element_text(size = 14),  # Enlarge y-axis label
    legend.position = "none"  # Remove legend to reduce clutter
  )

print(plot_county)

```

Then we analyze the geographic distribution of food scrap drop-off sites across New York City's counties. The plot indicates that some boroughs, such as Suffolk County and Westchester County, have notably higher counts. In contrast, others, like Cortland County and Seneca County, exhibit minimal values.



```{r}
# Load necessary libraries
library(leaflet)

# Filter rodent activity data for August
rodent_august <- data_rodent %>% filter(month == 8)

leaflet() %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  # Add rodent activity points
  addCircleMarkers(
    data = rodent_august,
    lng = ~longitude, lat = ~latitude,
    fillColor = "red",  # Fill color for rodent activity
    color = NA,  # No outline
    popup = ~paste("Rodent Activity:", result),
    radius = 4,
    fillOpacity = 0.7,
    group = "Rodent Activity"
  ) %>%
  # Add food scrap drop-off points
  addCircleMarkers(
    data = food_scrap_df,
    lng = ~longitude, lat = ~latitude,
    fillColor = "green",  # Fill color for food scrap sites
    color = NA,  # No outline
    popup = ~paste("Borough:", borough, "<br>Operation Type:", operation_type),
    radius = 4,
    fillOpacity = 0.7,
    group = "Food Scrap Sites"
  ) %>%
  # Add a layer control
  addLayersControl(
    overlayGroups = c("Rodent Activity", "Food Scrap Sites"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  # Add a legend
  addLegend(
    "bottomright",
    colors = c("red", "green"),
    labels = c("Rodent Activity", "Food Scrap Sites"),
    title = "Legend"
  )

```

These two graphs overlay rodent activity data and food scrap drop-off sites for August, a peak month for rodent activity. The visualization aims to understand the spatial relationship between these two factors.<br>
<br>
We found that rodent activity is heavily concentrated in high-density urban areas, particularly in Manhattan and parts of Brooklyn. Sparse activity is observed in Staten Island, which aligns with its lower population density.
<br>
As for their interaction, the distribution of food scrap sites shows a notable overlap with high rodent activity areas, particularly in Manhattan and Brooklyn. Staten Island has both fewer rodent activity reports and fewer food scrap sites, indicating a potential inequity in access to waste management resources.<br>





